Factors determining how catchments respond to forest cover change. Re-analysing global data sets.

### 2022-02-02 paper/Forest_and_Water/Forest_and_Water.Rmd
updates paper 02022022

In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.

@@

<!-- read in the ET data -->

```{r, echo = F, warning = F, message=F}
PET_large <- read_csv("../../data/largeET.csv")
PET_small <- read_csv("../../data/smallET.csv")

@@

# 
PET_all <- bind_rows(PET_large, PET_small)
# There are 7 missing E0 values
# > PET_all$`Watershed #`[is.na(PET_all$E0)==T]
# [1]  38 278 269 243 236 299 250
```

@@

<!--now combine with E0 values -->
```{r, echo = F}
All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
All_data <- All_data %>%
  mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
  select(-c(E0.x,E0.y)) #%>%
  #rename(Latitude = Latitude.x,
#         Longitude = Longitude.x)

@@

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

@@

```{r, echo = F}
All_data2 <- All_data %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "decrease", "increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
```

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest change direction", 
                      values = c("increase" = "blue", "decrease" = "red")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

@@

include_graphics("../../data/FAOET0data2.png")
```

@@

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
```

The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change. In this case the evidence is not very strong (p = `r round(summary(model4_dryness)$p.table[4,4],2)`. However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is ~6\% greater. Again, overall variance explained is not very much, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.

@@

There possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution. These catchments are listed in Table \@ref(tab:drytable).

@@

The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017. Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.

@@

The results of this model (Equation \@ref(eq:eq6)) clearly indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)). In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by 3.5\%. Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area. Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).

@@

<!--filter length < 60 -->
```{r model7_nolongLength, echo = F}
NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = NoLong_Data)
#summary(model7)
```
```{r, echo = F}
pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
```
```{r visualisesmooths7,echo =F, eval = F}
#require(visreg)
png("model7_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
(D + A) / (L + From) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```
```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}

@@

>>>>

In addition, additional variables added were the latitude and longitude for the centre of the catchment as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018]. For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017 and the additional studies, 29 potential duplicates were identified and flagged in the data.

@@

<!-- read in the ET data -->
<!-- no longer needed as this is now included in the data
```{r, echo = F, warning = F, message=F}
PET_large <- read_csv("../../data/largeET.csv")
PET_small <- read_csv("../../data/smallET.csv")

@@

# 
PET_all <- bind_rows(PET_large, PET_small)
# There are 7 missing E0 values
# > PET_all$`Watershed #`[is.na(PET_all$E0)==T]
# [1]  38 278 269 243 236 299 250
```
--->

@@

```{r, eval = F, echo = F}
# checking missing values
summary(All_data)

<!--now combine with E0 values -->
<!-- No longer needed as now included in the data -->
```{r, echo = F}
# All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
# All_data <- All_data %>%
#   mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
#   select(-c(E0.x,E0.y)) #%>%
#   #rename(Latitude = Latitude.x,
# #         Longitude = Longitude.x)

@@

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual E0 (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

@@

```{r, echo = F}
All_data2 <- All_data %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "Decrease", "Increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
```

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  filter(is.na(DeltaF_perc_pos) == F) %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest change direction", 
                      values = c("Decrease" = "red", "Increase" = "blue")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

@@

include_graphics("../../data/FAOET0data.png")
```

@@

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
```
```{r testing_dryness, echo = F, eval= F}
All_data2 %>% ggplot(aes(Dryness,log10(Area_km2), colour = DeltaF_perc_pos)) + geom_point() + theme_bw()
```
```{r testing_PCA, eval = F, echo = F, warning = F}
require(FactoMineR)
require(factoextra)
 res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
 # scree.plot <- fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 41))
 #  scree.plot
 #  
 #   var.plot <- fviz_pca_var(res.pca, axes = c(1, 2), col.var = "black")
 #  var.plot
 png("PCA_biplot_Alldata.png", width = 800, height = 800)
  biplot_test <- fviz_pca_biplot(res.pca, axes = c(1, 2), repel = TRUE,
                                                      col.var = "darkblue", # Variables color
                            label="var", #"var" 
                            col.ind = "red", # Individuals color
                            title = "",labelsize = 8, pointsize = 4, arrowsize = 1) +
        theme(text = element_text(size = 14, face = "bold"),
              axis.title = element_text(size = rel(2), face = "bold"),
        axis.text = element_text(size = rel(2), face = "bold"))
 print(biplot_test)
 dev.off()

The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results [@zhang2017; @filoso2017]. In this case the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`). However, it is very well possible that there is a further interaction in the data that this simpler version of the model cannot identify. Again, overall variance explained is not very much, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.

@@

There are also possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution. These catchments are listed in Table \@ref(tab:drytable).

@@

The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017. Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*. The effect of Dryness remains in the model, given that there could be interactions that are worth exploring.

@@

The results of this model (Equation \@ref(eq:eq6)) indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)). In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by 3.5\%. Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area. Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).

@@

<!--filter length < 60 -->
```{r model7_nolongLength, echo = F}
NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1930)
model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = NoLong_Data)
#summary(model7)
```
```{r, echo = F}
pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
```
```{r visualisesmooths7,echo =F, eval = F}
#require(visreg)
png("model7_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
(D + A) / (L + From) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```
```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model with reduced data for dryness and length", out.width="90%", echo=F}

@@

```{r model9_interact, echo = F, eval = F}
All_data3 <- All_data2 %>%
  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )

model9_interact <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
                      ti(Dryness,From, bs="ts" ) + #s(Latitude, bs="ts") + 
                     # ti(log10(Area_km2),From, bs="ts" ) +
                     # log10(Area_km2):Assessment_technique +
                      s(length, bs = "ts") + 
                      s(From, bs = "ts") +
                    Precip_data_type + Forest_type + Assessment_technique +
                    Hydrological_regime, data = All_data3 %>% filter(length < 60, Dryness <= 4))
#gam.check(model9_interact)
# plot(model9_interact)
summary(model9_interact)
